Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-25 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{19 February 2017}

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding\textsuperscript{1--3}. In response to this, genetic
instrumentation has emerged as a technique for improving the reliability
of causal inference in observational data, and with the coincident rise
in genome-wide association studies it is now a prominent tool that is
applied in several different guises\textsuperscript{3--6}. However,
potential pitfalls remain and one that is often neglected is the
influence of non-differential measurement error on the reliability of
causal inference.

Measurement error is the difference between the measured value of a
quantity and its true value. This study focuses specifically on
non-differential measurement error where all strata of a measured
variable have the same error rate, which can manifest as changes in
scale or levels of imprecision (noise) of the measurement. Such
variability can arise through a whole plethora of mechanisms, which are
often unique to the study design and difficult to
avoid\textsuperscript{7,8}. Array technology is now commonly used to
obtain high throughput phenotyping at low cost, but comes with the
problem of having imperfect resolution, for instance methylation levels
as measured by the Illumina450k chip are prone to have some amount of
noise around the true value due to imperfect
sensitivity\textsuperscript{9,10}. Relatedly, if the measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types\textsuperscript{11}.

Measurement error will of course arise in other types of data too. For
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13,14}. Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

An established study design that can provide information about causality
is randomisation. Given the hypothesis that trait A (henceforth referred
to as the exposure) is causally related to trait B (henceforth referred
to as the outcome), randomisation can be employed to assess the causal
nature of the association by randomly splitting the sample into two
groups, subjecting one group to the exposure and treating the other as a
control. The association between the exposure and the outcome in this
setting provides a robust estimate of the causal relationship. This
provides the theoretical basis behind randomised control trials, but in
practice randomisation is often impossible to implement in an
experimental context due to cost, scale or inability to manipulate the
exposure. The principle, however, can be employed in extant
observational data through the use of genetic variants associated with
the exposure (instruments), where the inheritance of an allele serves as
a random lifetime allocation of differential exposure
levels\textsuperscript{15,16}. Two statistical approaches to exploiting
the properties of genetic instruments are widely used: mediation-based
approaches and Mendelian randomisation (MR).

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that the indirect influence of the SNP on the outcome will
disappear when conditioning on the exposure. Here, the exposure
`mediates' the association between the SNP and the outcome, providing
information about the causal influence of the exposure on the outcome.
This forms the basis of a number of methods such as genetical
genomics\textsuperscript{17}, the regression-based causal inference test
(CIT)\textsuperscript{4,18}, a structural equation modelling (SEM)
implementation in the NEO software\textsuperscript{5}, and various other
methods including Bayesian approaches\textsuperscript{6}. They have been
employed by a number of recent publications that make causal inferences
in large scale `omics datasets\textsuperscript{6,19--23}.

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP is used as a surrogate for the exposure. Assuming the SNP
associates with the outcome only through the exposure, the causal effect
of the exposure on the outcome can be estimated by scaling the
association between the SNP and the outcome by the association between
the SNP and the exposure. Though difficult to test empirically, this
assumption can be relaxed in various ways when multiple instruments are
available for a putative exposure\textsuperscript{24,25} and a number of
sensitivity tests are now available to improve
reliability\textsuperscript{26}.

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. In the CIT framework
(described fully in the Methods) for example, the test statistic is
different if you test for the exposure causing the outcome or the
outcome causing the exposure, allowing the researcher to infer the
direction of causality between two variables by performing the test in
both directions and choosing the model with the strongest evidence. The
CIT also has the valuable property of being able to distinguish between
several putative causal graphs that link the traits with the SNP (Figure
1). Such is not the case for MR, where in order to infer the direction
of causality between two traits the instrument must be known to
influence the exposure primarily, associating with the outcome only
through the exposure.

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the SNP being robustly associated with each, it can be difficult to
determine which of the two variables is subject to the primary effect of
the SNP (i.e.~for which of the two variables is the SNP a valid
instrument? Figure 1). By definition, we expect that if the association
is causal then a SNP for the exposure will be associated with the
outcome, such that if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. Genome-wide association studies (GWASs) that
identify genetic associations for complex traits are, by design,
hypothesis free and agnostic of genomic function, and it often takes
years of follow up studies to understand the biological nature of a
putative GWAS hit\textsuperscript{27}. If multiple instruments are
available for an hypothesised exposure, which is increasingly typical
for complex traits that are analysed in large GWAS consortia, then
techniques can be applied to mitigate these issues\textsuperscript{16}.
But these techniques cannot always be applied in the case of determining
causal directions between 'omic measures where typically only one
cis-acting SNP is known. For example if a DNA methylation probe is
associated with expression of an adjacent gene, then is a cis-acting SNP
an instrument for the DNA methylation level, or the gene expression
level (Figure 1)?

MR has some important advantages over the mediation-based approaches.
The mediation-based approaches require that the exposure, outcome and
instrumental variables are all measured in the same data, whereas recent
extensions to MR circumvent this requirement, allowing causal inference
to be drawn when exposure variables and outcome variables are measured
in different samples\textsuperscript{28}. This has the crucial advantage
of improving statistical power by allowing analysis in much larger
sample sizes, and dramatically expands the breadth of possible
phenotypic relationships that can be evaluated\textsuperscript{26}.
Additionally, when MR assumptions are satisfied the method is robust to
there being measurement error in the exposure
variable\textsuperscript{29}. Indeed instrumental variable (IV) analysis
was in part initially introduced as a correction for measurement error
in the exposure\textsuperscript{30}, whereas it has been noted that both
classic mediation-based analyses\textsuperscript{13,14,31,32} and
mediation-based methods that use instrumental
variables\textsuperscript{33,34} are prone to be unreliable in its
presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. We then present an extension to MR that
allows researchers to ascertain the causal direction of an association
even when the biology of the instruments are not fully understood, and
also a matric to evaluate the sensitivity of the result of this
extension to measurement error. Together these extensions improve the
utility of MR in cases where mediation based methods might have
otherwise been used preferentially. Finally, we apply this new method to
infer the direction of causation between DNA methylation levels and gene
expression levels. Our analyses highlight that because these different
causal inference techniques have varying strengths and weaknesses,
triangulation of evidence from as many sources as possible should be
practiced in causal inference\textsuperscript{35}.

@@

Measurement error of an exposure can be modeled as a transformation of
the true value that leads to the observed value, \(x_o = f(x)\). For
example, following Pierce and VanderWeele\textsuperscript{29} we can
define

@@

The CIT method\textsuperscript{4} is implemented in the R package
\emph{R/cit}\textsuperscript{18}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) causes an outcome \(y\), as described above. The
following tests are then performed:

@@

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
for the direction of \(x\) causing \(y\) (Model 1), and for the
direction of \(y\) causing \(x\) (Model 2). For some significance
threshold \(\alpha\), the existence of causality and its direction was
inferred based on the following scenarios:

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance
thresholds\textsuperscript{36}.

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the R package \(R/systemfit\)\textsuperscript{37}.

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{38}. It is calculated as

@@

and a p-value is generated from the \(Z\) value to indicate the
probability of obtaining the observed difference in correlations
\(\rho_{gx}\) and \(\rho_{gy}\) under the null hypothesis that both
correlations are identical. The existence of causality and its direction
is inferred based on the following scenarios:

@@

For the purposes of compiling simulation results, we use an arbitrary
\(\alpha = 0.05\) value.

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{28} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (e.g.~from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{39}.

@@

Two sample MR\textsuperscript{28} was performed using the summary
statistics for genetic influences on gene expression and DNA
methylation. To do this we obtained a list of 458 gene expression - DNA
methylation associations as reported in Shakhbazov et
al\textsuperscript{40}. These were filtered to be located on the same
chromosome, have robust correlations after correcting for multiple
testing, and to share a SNP that had a robust cis-acting effect on both
the DNA methylation probe and the gene expression probe. Because only
summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming
language\textsuperscript{41} and code is made available at
\href{}{github url to go here} and implemented in the MR-Base
(\url{http://wwww.mrbase.org}) platform\textsuperscript{26}.

@@

We performed two sample MR\textsuperscript{28} for each association in
the direction of causality inferred by the Stieger test. We observed
that the sign of the MR estimate was generally in the same direction as
the Pearson correlation coefficient reported by Shakhbazov et
al\textsuperscript{40} (Figure 6b). There was a moderate correlation
between the absolute magnitudes of the causal correlation and the
observational Pearson correlation (r = 0.45). Together these inferences
suggest that even for relatively low level associations it is important
to use causal inference methods over observational associations to infer
causal effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing gene expression
levels at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation\textsuperscript{42}.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored\textsuperscript{43}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of methods such as the CIT or similar
that are based on mediation. Inferring the existence of causality using
Mendelian randomisation is robust in the face of measurement error and,
if the researcher has knowledge about the biology of the instrument
being used in the analysis, can offer a direct solution to the issues
that the CIT faces. This assumption is often reasonable, for example
SNPs are commonly used as instruments when they are found in genes with
known biological relevance for the trait of interest. But on many
occasions, especially in the realm of 'omic data, this is not the case,
and methods based on mediation have been valuable in order to be able to
both ascertain if there is a causal association and to infer the
direction of causality. Here we have described a simple extension to MR
which can be used as an alternative to or in conjunction with mediation
based methods. We show that this method is still liable to measurement
error, but because it has different properties to the CIT it offers
several main advantages. First, it uses a formal statistical framework
to test for the robustness of the assumed direction of causality.
Second, after testing in a comprehensive range of scenarios the MR based
approach is less likely to infer the wrong direction of causality
compared to CIT, while substantially improving power over CIT in the
cases where \(d > 0\).

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this
position\textsuperscript{44,45}, though making strong conclusions for
this analysis is difficult because measurement error is likely to be
study specific.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
MR strategies exist for other scenarios, for example if there are
multiple instruments for the hypothesised exposure then the problems
described here may be circumvented\textsuperscript{16}. Likewise, if
there is at least one instrument for each trait then bi-directional MR
can offer solutions to inferring the causal
direction\textsuperscript{46}. We restricted the simulations to
evaluating the causal inference between quantitative traits, but it is
possible that the analysis could be extended to binary traits by using
the genetic variance explained on the liability scale, taking into
account the population prevalence\textsuperscript{47}.

In this work we assumed that pleiotropy (the influence of the instrument
on the outcome through a mechanism other than the exposure) was not
present. Recent method developments in MR\textsuperscript{24,25} have
focused on accounting for the issues that horizontal pleiotropy can
introduce when multiple instruments are available, but how they perform
in the presence of measurement error remains to be explored. An
important advantage that MR confers over most mediation based analysis
is that it can be performed in two samples, which can considerably
improve power and expand the scope of analysis. However, whether there
is a substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established\textsuperscript{34} and have a
number of extensions that make them valuable tools, including for
example network construction. But because they are predicated on the
basic underlying principles of mediation they are liable to suffer from
the same issues of measurement error. Recent advances in MR methodology,
for example applying MR to genetical genomics\textsuperscript{48},
multivariate MR\textsuperscript{49} and mediation through
MR\textsuperscript{50--52} may offer more robust alternatives for these
more complicated problems.

@@

@@

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-Phillips1991}{}
1. Phillips, A. N. \& Davey Smith, G. How independent are `independent'
effects? relative risk estimation when correlated exposures are measured
imprecisely. \emph{Journal of Clinical Epidemiology} \textbf{44,}
1223--1231 (1991).

\hypertarget{ref-DaveySmith2002}{}
2. Davey Smith, G. \& Ebrahim, S. Data dredging, bias, or confounding.
\emph{BMJ} \textbf{325,} (2002).

\hypertarget{ref-DaveySmith2004}{}
3. Davey Smith, G. \& Ebrahim, S. Mendelian randomization: prospects,
potentials, and limitations. \emph{International journal of
epidemiology} \textbf{33,} 30--42 (2004).

>>>>

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{24 February 2017}

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding (1--3). In response to this, genetic instrumentation has
emerged as a technique for improving the reliability of causal inference
in observational data, and with the coincident rise in genome-wide
association studies it is now a prominent tool that is applied in
several different guises (3--6). However, potential pitfalls remain and
one that is often neglected is the influence of non-differential
measurement error on the reliability of causal inference.

Measurement error is the difference between the measured value of a
quantity and its true value. This study focuses specifically on
non-differential measurement error where all strata of a measured
variable have the same error rate, which can manifest as changes in
scale or levels of imprecision (noise) of the measurement. Such
variability can arise through a whole plethora of mechanisms, which are
often unique to the study design and difficult to avoid (7,8). Array
technology is now commonly used to obtain high throughput phenotyping at
low cost, but comes with the problem of having imperfect resolution, for
instance methylation levels as measured by the Illumina450k chip are
prone to have some amount of noise around the true value due to
imperfect sensitivity (9,10). Relatedly, if the measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types (11).

Measurement error will of course arise in other types of data too. For
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect (12). A similar problem of
biological misspecification is unavoidable in disease diagnosis, and
measuring behaviour such as smoking or diet is notoriously difficult to
do accurately. Measurement error can also be introduced after the data
have been collected, for example the transformation of non-normal data
for the purpose of statistical analysis will lead to a new variable that
will typically incur both changes in scale and imprecision (noise)
compared to the original variable. The sources of measurement error are
not limited to this list (8), and its impact has been explored in the
epidemiological literature extensively (13,14). Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

An established study design that can provide information about causality
is randomisation. Given the hypothesis that trait A (henceforth referred
to as the exposure) is causally related to trait B (henceforth referred
to as the outcome), randomisation can be employed to assess the causal
nature of the association by randomly splitting the sample into two
groups, subjecting one group to the exposure and treating the other as a
control. The association between the exposure and the outcome in this
setting provides a robust estimate of the causal relationship. This
provides the theoretical basis behind randomised control trials, but in
practice randomisation is often difficult or impossible to implement in
an experimental context due to cost, scale or inability to manipulate
the exposure. The principle, however, can be employed in extant
observational data through the use of genetic variants associated with
the exposure (instruments), where the inheritance of an allele serves as
a random lifetime allocation of differential exposure levels (15,16).
Two statistical approaches to exploiting the properties of genetic
instruments are widely used: mediation-based approaches and Mendelian
randomisation (MR).

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that the indirect influence of the SNP on the outcome will
disappear when conditioning on the exposure. Here, the exposure
`mediates' the association between the SNP and the outcome, providing
information about the causal influence of the exposure on the outcome.
This forms the basis of a number of methods such as genetical genomics
(17), the regression-based causal inference test (CIT) (4,18), a
structural equation modelling (SEM) implementation in the NEO software
(5), and various other methods including Bayesian approaches (6). They
have been employed by a number of recent publications that make causal
inferences in large scale `omics datasets (6,19--23).

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP is used as a surrogate for the exposure. Assuming the SNP
associates with the outcome only through the exposure, the causal effect
of the exposure on the outcome can be estimated by scaling the
association between the SNP and the outcome by the association between
the SNP and the exposure. Though difficult to test empirically, this
assumption can be relaxed in various ways when multiple instruments are
available for a putative exposure (24,25) and a number of sensitivity
tests are now available to improve reliability (26).

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. In the CIT framework
(described fully in the Methods) for example, the test statistic is
different if you test for the exposure causing the outcome or the
outcome causing the exposure, allowing the researcher to infer the
direction of causality between two variables by performing the test in
both directions and choosing the model with the strongest evidence. The
CIT also has the valuable property of being able to distinguish between
several putative causal graphs that link the traits with the SNP (Figure
1). Such is not the case for MR, where in order to infer the direction
of causality between two traits the instrument must have its most
proximal link with the exposure, associating with the outcome only
through the exposure.

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the SNP being robustly associated with each, it can be difficult to
determine which of the two variables is subject to the primary effect of
the SNP (i.e.~for which of the two variables is the SNP a valid
instrument? Figure 1). By definition, we expect that if the association
is causal then a SNP for the exposure will be associated with the
outcome, such that if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. Genome-wide association studies (GWASs) that
identify genetic associations for complex traits are, by design,
hypothesis free and agnostic of genomic function, and it often takes
years of follow up studies to understand the biological nature of a
putative GWAS hit (27). If multiple instruments are available for an
hypothesised exposure, which is increasingly typical for complex traits
that are analysed in large GWAS consortia, then techniques can be
applied to mitigate these issues (16). But these techniques cannot
always be applied in the case of determining causal directions between
'omic measures where typically only one cis-acting SNP is known. For
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level (Figure 1)?

MR has some important advantages over the mediation-based approaches.
The mediation-based approaches require that the exposure, outcome and
instrumental variables are all measured in the same data, whereas recent
extensions to MR circumvent this requirement, allowing causal inference
to be drawn when exposure variables and outcome variables are measured
in different samples (28). This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes, and
dramatically expands the breadth of possible phenotypic relationships
that can be evaluated (26). Additionally, when MR assumptions are
satisfied the method is robust to there being measurement error in the
exposure variable (29). Indeed instrumental variable (IV) analysis was
in part initially introduced as a correction for measurement error in
the exposure (30), whereas it has been noted that both classic
mediation-based analyses (13,14,31,32) and mediation-based methods that
use instrumental variables (33,34) are prone to be unreliable in its
presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. We then present an extension to MR that
allows researchers to ascertain the causal direction of an association
even when the biology of the instruments are not fully understood, and
also a metric to evaluate the sensitivity of the result of this
extension to measurement error. Together these extensions improve the
utility of MR in cases where mediation based methods might have
otherwise been used preferentially. Finally, we apply this method to
infer the direction of causation between DNA methylation levels and gene
expression levels. Our analyses highlight that because these different
causal inference techniques have varying strengths and weaknesses,
triangulation of evidence from as many sources as possible should be
practiced in causal inference (35).

@@

Measurement error of an exposure can be modeled as a transformation of
the true value that leads to the observed value, \(x_o = f(x)\). For
example, following Pierce and VanderWeele (29) we can define

@@

The CIT method (4) is implemented in the R package \emph{R/cit} (18).
The methodology of the CIT is as follows. Assume an exposure \(x\) is
instrumented by a SNP \(g\), and the exposure \(x\) causes an outcome
\(y\), as described above. The following tests are then performed:

@@

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT for each of two tests -
\(p_{CIT, x \rightarrow y}\), was estimated for the direction of \(x\)
causing \(y\) (Model 1), and for the direction of \(y\) causing \(x\),
\(p_{CIT, y \rightarrow x}\) (Model 2). The results from each of these
methods can then be used in combination to infer the existance and
direction of causality. For some significance threshold \(\alpha\) there
are four possible outcomes from these two tests, and their
interpretations are as follows:

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds (36).

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the R package \(R/systemfit\) (37).

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population (38). It is calculated as

@@

and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
indicate the probability of obtaining the observed difference in
correlations \(\rho_{gx}\) and \(\rho_{gy}\) under the null hypothesis
that both correlations are identical.

The existence of causality and its direction is inferred based on
combining information from the MR analysis and the Steiger test. The MR
analysis indicates whether there is a potential causal relationship
(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
of the causal relationship and the confidence of the direction
(\$p\_\{Steiger\}). For the purposes of compiling simulation results,
these can be combined using an arbitrary \(\alpha = 0.05\) value:

@@

Note that the same correlation test approach can be applied to a
two-sample MR (28) setting. Two-sample MR refers to the case where the
SNP-exposure association and SNP-outcome association are calculated in
different samples (e.g.~from publicly available summary statistics).
Here the Steiger test of two independent correlations can be applied
where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package \emph{R/psych} (39).

@@

Two sample MR (28) was performed using the summary statistics for
genetic influences on gene expression and DNA methylation. To do this we
obtained a list of 458 gene expression - DNA methylation associations as
reported in Shakhbazov et al (40). These were filtered to be located on
the same chromosome, have robust correlations after correcting for
multiple testing, and to share a SNP that had a robust cis-acting effect
on both the DNA methylation probe and the gene expression probe. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming language (41) and
code is made available at \href{}{github url to go here} and implemented
in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).

@@

We performed two sample MR (28) for each association in the direction of
causality inferred by the Stieger test. We observed that the sign of the
MR estimate was generally in the same direction as the Pearson
correlation coefficient reported by Shakhbazov et al (40) (Figure 6b).
There was a moderate correlation between the absolute magnitudes of the
causal correlation and the observational Pearson correlation (r = 0.45).
Together these inferences suggest that even for relatively low level
associations it is important to use causal inference methods over
observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing methylation levels
at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation (42).

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored (43). Our analysis extends this to
related methods such as CIT that are widely used in predominantly 'omic
data. These methods are indeed liable to the same issues as standard
mediation based analysis, and specifically we show that as measurement
error in the exposure variable increases, CIT is likely to have reduced
statistical power, and liable to infer the wrong direction of causality.
We also demonstrate that, though unintuitive, increasing sample size
does not resolve the issue, rather it leads to more extreme p-values for
the model that predicts the wrong direction of causality.

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of methods such as the CIT or similar
that are based on mediation. Inferring the existence of causality using
Mendelian randomisation is robust in the face of measurement error and,
if the researcher has knowledge about the biology of the instrument
being used in the analysis, can offer a direct solution to the issues
that the CIT faces. This assumption is often reasonable, for example
SNPs are commonly used as instruments when they are found in genes with
known biological relevance for the trait of interest. But on many
occasions, especially in the realm of 'omic data, this is not the case,
and methods based on mediation have been valuable in order to be able to
both ascertain if there is a causal association and to infer the
direction of causality. Here we have described a simple extension to MR
which can be used as an alternative to or in conjunction with mediation
based methods. We show that this method is still liable to measurement
error, but because it has different properties to the CIT it offers
several main advantages. First, it uses a formal statistical framework
to test for the reliability of the assumed direction of causality.
Second, after testing in a comprehensive range of scenarios the MR based
approach is less likely to infer the wrong direction of causality
compared to CIT, while substantially improving power over CIT in the
cases where \(d > 0\).

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this position
(44,45), though making strong conclusions for this analysis is difficult
because measurement error is likely to be study specific.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. We note
that it is straightforward to extend the MR Steiger approach to multiple
instruments, requiring only that the total variance explained by all
instruments be calculated under the assumption that they are
independent. Multiple instruments can indeed help to circumvent some of
the problems described here in other ways also, for example by
evaluating the proportionality of the SNP-exposure and SNP-outcome
effects (16). Additionally, if there is at least one instrument for each
trait then bi-directional MR can offer solutions to inferring the causal
direction (46). We restricted the simulations to evaluating the causal
inference between quantitative traits, but it is possible that the
analysis could be extended to binary traits by using the genetic
variance explained on the liability scale, taking into account the
population prevalence (47).

In this work we assumed that pleiotropy (the influence of the instrument
on the outcome through a mechanism other than the exposure) was not
present. Recent method developments in MR (24,25) have focused on
accounting for the issues that horizontal pleiotropy can introduce when
multiple instruments are available, but how they perform in the presence
of measurement error remains to be explored. An important advantage that
MR confers over most mediation based analysis is that it can be
performed in two samples, which can considerably improve power and
expand the scope of analysis. However, whether there is a substantive
difference in two sample MR versus one sample MR in how measurement
error has an effect is not yet fully understood. We have also assumed no
measurement error in the genetic instrument, which is not unreasonable
given the strict QC protocols that ensure high quality genotype data is
available to most studies. We have restricted the scope to only
exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated. We have also not addressed other issues pertaining to
instrumental variables, such as exposure missspecification, where the
instrument is used for a particular variable due to that variable's
non-causal correlation with the true exposure. However our analysis goes
beyond many previous explorations of measurement error by assessing the
impacts of both imprecision (noise) and linear transformations of the
true variable on causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established (34) and have a number of
extensions that make them valuable tools, including for example network
construction. But because they are predicated on the basic underlying
principles of mediation they are liable to suffer from the same issues
of measurement error. Recent advances in MR methodology, for example
applying MR to genetical genomics (48), multivariate MR (49) and
mediation through MR (50--52) may offer more robust alternatives for
these more complicated problems.

@@

\begin{itemize}
\tightlist
\item
  other types of measurement error, e.g.~differential; exposure
  missspecification;
\item
  using multiple instruments
\end{itemize}

@@

\subsection{References}\label{references}

\raggedright

\hypertarget{refs}{}
\hypertarget{ref-Phillips1991}{}
1. Phillips AN, Davey Smith G. How independent are ``independent''
effects? relative risk estimation when correlated exposures are measured
imprecisely. Journal of Clinical Epidemiology. 1991;44(11):1223--31.

\hypertarget{ref-DaveySmith2002}{}
2. Davey Smith G, Ebrahim S. Data dredging, bias, or confounding. BMJ.
2002;325(7378):1437--8.

\hypertarget{ref-DaveySmith2004}{}
3. Davey Smith G, Ebrahim S. Mendelian randomization: prospects,
potentials, and limitations. International journal of epidemiology
{[}Internet{]}. 2004 Feb;33(1):30--42. Available from:
\url{http://www.ncbi.nlm.nih.gov/pubmed/15075143}
